---
title: "Project Draft"
author: "Luke Fredrickson, Anika Miner, Chris McCabe"
date: "4/9/2021"
output: 
        html_document: 
                toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r import,message = FALSE,warning=FALSE}

library(tidyverse)
library(ggmap)
library(viridis)

properties <- read.csv("properties.csv")


```

## Number of Properties

Out of the 3086 entries in the rental property data we obtained, 2121 of them are
owned by owners who own less than 5 properties. Most of that 2121 is made up of
the 1809 properties whose owners only own that one property. So to take a deeper
look at some of the major landlords and property management companies in Burlington,
we filtered the data to include properties whose owners owned at least 5 properties.
Shown in the second graph, you can see that a majority of the owners that own more than
four properties own 5-7 properties. The outlier in the data set is Diemer Properties
who owns a total of 32 properties. 

```{r num_properties}

NumProperties <- properties %>% group_by(Owner) %>% summarize(nProperties = n())

summary(NumProperties$nProperties)

ggplot(data = NumProperties, 
       mapping = aes(nProperties)) + 
        geom_bar(color = "black",fill="tomato2")+
        labs(title = "Properties Owned per Landlord",x="Number of Properties Owned",
             y = "Owners")+
        theme_minimal()

ggplot(data = NumProperties %>% filter(nProperties >= 5),mapping = aes(nProperties)) +
        geom_bar(color = "black",fill="tomato2")+
        labs(title = "Properties Owned per Landlord",x="Number of Properties Owned",
             y = "Owners",subtitle = "Owners With 5 or More Properties")+
         annotate( geom = 'text', x = 31, y = 2, label = "Diemer Properties")+
        theme_minimal()
```

## Types of Rental Properites

There are 15 active land use codes in our dataset for the rental properties in Burlington.
The most common is a two family rental, with one to 5 family apartments, and residential 
condos taking up the very large majority of properties. The commerical/residential 
properties (residential above commercial properties e.g. Church Street) are the next
most common.

```{r rental_types}

ggplot(data = properties %>% filter(!is.na(LandUseCode)),mapping = aes(x = LandUseCode,fill = LandUseCode)) + 
        geom_bar()+
        labs(title = "Rental Property Land Use Codes", 
             x = "Land Use Code", 
             y = "Count",
             fill = "Land Use Codes")+
        theme_minimal()+
        scale_fill_discrete(labels = c("Commercial","Commericial Condo",
                                       "Commercial/Residential","Exempt(Land+Building)","Mobile Home(w/o land)",
                                       "Single Family","Two Family","Three Family","Four Family","Five+ Family",
                                       "Res. Apt. Condo","Residential Condo","Service","Taxable/Partly Exempt",
                                       "Unknown Owner"))

```

## Certificate of Compliance

The certificate of compliance is based off house inspections and how many years
in a row the property has complied with Burlington's building code. If the 
property has complied with the code for 5 years, they will have a 5 year Certificate
of Compliance. The majority of properties have 4 or 5 year certificates. Properties
with a 0 year certificate either failed inspections in the most recent inspection
or have not been inspected due to COVID-19.


```{r CoC_type}

ggplot(data = properties %>% filter(!is.na(CoCYears)), mapping = aes(x=CoCYears)) + 
        geom_bar(color = "black",fill = "#CE2029")+
        labs(title = "Certificate of Compliance Age",
             subtitle = "(Higher is Better)",
             x="Certificate of Compliance Age (Years)", 
             y = "Number of Properties")+
        theme_minimal()

```

## Does CoC Affect Property Value

The original hypothesis was that the properties with a less desirable CoC would
have a lower property value. However, from the boxplots below we learned that the
property value across all year values for the CoC was very similar. The median 
property value for all CoC years was ~$275,000.


```{r value_vs_coc}

ggplot(data = properties %>% 
               filter(!is.na(CoCYears) & CoCYears != 0 & !is.na(LastMhInspectionDate) & PropertyValue < 1000000), 
       mapping = aes(x=factor(CoCYears), y=PropertyValue,fill = CoCYears)) + 
        geom_boxplot()+
        labs(title = "Property Value VS. CoC Age",
             x = "Certificate of Compliance Age (Years)",
             y = "Property Value (USD)")+
        scale_fill_gradient(high = "#ff9300",low = "#b60000")+
        guides(fill=FALSE)+
        theme_minimal()


```


```{r top_20_landlords}

properties %>%
  group_by(Owner) %>%
  summarize(
    NumberOfProperties = n(),
    TotalValue = sum(PropertyValue)
  ) %>% 
  arrange(desc(NumberOfProperties)) %>% 
  top_n(20, NumberOfProperties)

```